### 2023-07-25 tex/agda-primer.tex
[ writing ] full draft

Simple function types take the form
\AgdaArgument{A}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaArgument{B},
coinciding with Haskell's syntax.
Also as in Haskell and ML, the function arrow nests to the right.
However, Agda has a termination checker ensuring that all defineable functions
are total, so many Haskell functions do not have a corresponding Agda function.

@@

Agda uses $\Pi$-types where in Haskell we would use polymorphism.
For example, we can define an identity function as below.
The definition relies on quantifying over terms of type \AgdaPrimitive{Set},
i.e.\ (small) types.
This definition also gives an example of defining a function with an implicit
argument (\AgdaBound{X}), which can typically be inferred from either the
kargument type or the return type, so can be omitted.

@@

In the following definition \AgdaFunction{lemma}, we want to chase equations in
order to prove that \AgdaBound{x} is propositionally equal to \AgdaBound{z}.
We start with the following incomplete definition, where the expression
\AgdaHole{\{ \}0} marks an interation point, or \emph{hole}, in the program, to
which we can apply interactive commands to complete the program.

@@

Full \emph{dependent} pattern matching, as described by \citet{MM04}, is when
the unification of indices described above takes account of constructors.
In particular, the constructors of a data type satisfy the ``no confusion''
property --- constructors are injective and mutually disjoint.
Where we encounter disjoint constructors during unification, we may dismiss the
corresponding case as impossible.
Consider the following example (\AgdaFunction{elim-Fin-zero}).
We start with an argument
\AgdaBound{i}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaInductiveConstructor{zero}, and consider which constructors could possibly
construct such a value.
However, as noted earlier, both constructors of \AgdaDatatype{Fin} target
successor values of the index, from which \AgdaInductiveConstructor{zero} is
disjoint.
Therefore, both cases are impossible.
The notation when all cases are impossible is to place empty round brackets
\AgdaSymbol{()} in the place of the impossible argument, and to not provide a
righthand side to the clause.

@@

Ordinarily, each clause of a definition gives rise to a \emph{definitional}
equation between its lefthand side and righthand side.
In intensional type theory, as implemented by Agda, definitional and
propositional equality are contrasted to each other.
Definitional equality corresponds to a decidable fragment of the natural
equational theory of the type theory.
As such, definitional equality is an entirely metatheoretic notion, and we can
neither assume nor prove directly definitional equations within the language.
Definitional equality is sometimes also called \emph{judgemental equality},
because it forms a judgement which plays a part in the rules of the type theory.
As well as from the clauses of definitions, we also get definitional equations
from $\beta$-reductions of $\lambda$-abstractions and $\eta$-laws of functions
and records.
Because the type checker treats definitionally equal terms equivalently, we are
able to refactor up to definitional equality without changing any downstream
code.

@@

There are two main ways of using the fields of a record.
The first is to put the projections into scope using
\AgdaKeyword{open}\AgdaSpace{}\AgdaRecord{$\Upsigma$}, and then to use the field
names to project out of arbitrary terms of \AgdaRecord{$\Upsigma$}-type.
This is what I will always do when using the \AgdaRecord{$\Upsigma$}-type
family.
Within this paradigm, there are two further notational choices.
Either, we can use the field names as functions, so that
\AgdaBound{z}\AgdaSpace{}\AgdaSymbol{=}\AgdaSpace{}\AgdaField{proj$_1$}%
\AgdaSpace{}\AgdaBound{z}\AgdaSpace{}\AgdaOperator{\AgdaInductiveConstructor,}%
\AgdaSpace{}\AgdaField{proj$_2$}\AgdaSpace{}\AgdaBound{z},
or we can use postfix projections via the space-dot notation, as in
\AgdaBound{z}\AgdaSpace{}\AgdaSymbol{=}\AgdaSpace{}\AgdaBound{z}%
\AgdaSpace{}\AgdaSymbol.\AgdaField{proj$_1$}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor,}\AgdaSpace{}\AgdaBound{z}%
\AgdaSpace{}\AgdaSymbol.\AgdaField{proj$_2$}.
I tend to prefer the latter, also using it occasionally in ordinary mathematical
notation (without the space).
Both notations can also be used on the lefthand side of a clausal definition as
\emph{copatterns}.
Copatterns let us think of records as being function-like, with the fields of a
record type being the possible arguments we can pass to such a function.

@@

Along with \AgdaRecord{$\Upsigma$} and \AgdaRecord{$\top$}, there are two more
general-purpose record types I need to cover which take advantage of two special
features of \AgdaKeyword{record}-declarations (and also
\AgdaKeyword{data}-delcarations, but I use \AgdaKeyword{record}-declarations for
the convenience reason given in the previous paragraph).
The first feature is that the universe level of a record type has a lower bound
(the level of each field) but no upper bound.
Therefore, we can introduce the following declaration \AgdaRecord{Lift}, which
takes a type \AgdaBound{A} at level \AgdaBound{a} and produces an equivalent
type at a potentially higher level
\AgdaBound{a}\AgdaSpace{}\AgdaOperator{\AgdaPrimitive{$\sqcup$}}\AgdaSpace{}%
\AgdaBound{$\ell$}.
This type former is useful in situations which require a type at a specific
level, such as when constructing a type using a function.

>>>>

Simple function types take the form
\AgdaArgument{A}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaArgument{B},
coinciding with Haskell's syntax.
Also as in Haskell and ML, the function arrow nests to the right.
However, Agda has a termination checker ensuring that all definable functions
are total, so many Haskell functions do not have a corresponding Agda function.

@@

Agda uses $\Pi$-types where in Haskell we would use polymorphism.
For example, we can define an identity function as below.
The definition relies on quantifying over terms of type \AgdaPrimitive{Set},
i.e.\ (small) types.
This definition also gives an example of defining a function with an implicit
argument (\AgdaBound{X}), which can typically be inferred from either the
argument type or the return type, so can be omitted.

@@

In the following definition \AgdaFunction{lemma}, we want to chase equations in
order to prove that \AgdaBound{x} is propositionally equal to \AgdaBound{z}.
We start with the following incomplete definition, where the expression
\AgdaHole{\{ \}0} marks an interaction point, or \emph{hole}, in the program, to
which we can apply interactive commands to complete the program.

@@

Full \emph{dependent} pattern matching, as described by \citet{MM04}, is when
the unification of indices described above takes account of constructors.
In particular, the constructors of a data type satisfy the ``no confusion''
property --- constructors are injective and mutually disjoint.
Where we encounter disjoint constructors during unification, we may dismiss the
corresponding case as impossible.
Consider the following example (\AgdaFunction{elim-Fin-zero}).
We start with an argument
\AgdaBound{i}\AgdaSpace{}\AgdaSymbol:\AgdaSpace{}\AgdaDatatype{Fin}\AgdaSpace{}%
\AgdaInductiveConstructor{zero}, and consider which constructors could possibly
construct such a value.
However, as noted earlier, both constructors of \AgdaDatatype{Fin} target
successor values of the index, from which \AgdaInductiveConstructor{zero} is
disjoint.
Therefore, both cases are impossible.
The notation when all cases are impossible is to place empty round brackets
\AgdaSymbol{()} in the place of the impossible argument, and to not provide a
right-hand side to the clause.

@@

Ordinarily, each clause of a definition gives rise to a \emph{definitional}
equation between its left-hand side and right-hand side.
In intensional type theory, as implemented by Agda, definitional and
propositional equality are contrasted to each other.
Definitional equality corresponds to a decidable fragment of the natural
equational theory of the type theory.
As such, definitional equality is an entirely metatheoretic notion, and we can
neither assume nor prove directly definitional equations within the language.
Definitional equality is sometimes also called \emph{judgemental equality},
because it forms a judgement which plays a part in the rules of the type theory.
As well as from the clauses of definitions, we also get definitional equations
from $\beta$-reductions of $\lambda$-abstractions and $\eta$-laws of functions
and records.
Because the type checker treats definitionally equal terms equivalently, we are
able to refactor up to definitional equality without changing any downstream
code.

@@

There are two main ways of using the fields of a record.
The first is to put the projections into scope using
\AgdaKeyword{open}\AgdaSpace{}\AgdaRecord{$\Upsigma$}, and then to use the field
names to project out of arbitrary terms of \AgdaRecord{$\Upsigma$}-type.
This is what I will always do when using the \AgdaRecord{$\Upsigma$}-type
family.
Within this paradigm, there are two further notational choices.
Either, we can use the field names as functions, so that
\AgdaBound{z}\AgdaSpace{}\AgdaSymbol{=}\AgdaSpace{}\AgdaField{proj$_1$}%
\AgdaSpace{}\AgdaBound{z}\AgdaSpace{}\AgdaOperator{\AgdaInductiveConstructor,}%
\AgdaSpace{}\AgdaField{proj$_2$}\AgdaSpace{}\AgdaBound{z},
or we can use postfix projections via the space-dot notation, as in
\AgdaBound{z}\AgdaSpace{}\AgdaSymbol{=}\AgdaSpace{}\AgdaBound{z}%
\AgdaSpace{}\AgdaSymbol.\AgdaField{proj$_1$}\AgdaSpace{}%
\AgdaOperator{\AgdaInductiveConstructor,}\AgdaSpace{}\AgdaBound{z}%
\AgdaSpace{}\AgdaSymbol.\AgdaField{proj$_2$}.
I tend to prefer the latter, also using it occasionally in ordinary mathematical
notation (without the space).
Both notations can also be used on the left-hand side of a clausal definition as
\emph{copatterns}.
Copatterns let us think of records as being function-like, with the fields of a
record type being the possible arguments we can pass to such a function.

@@

Along with \AgdaRecord{$\Upsigma$} and \AgdaRecord{$\top$}, there are two more
general-purpose record types I need to cover which take advantage of two special
features of \AgdaKeyword{record}-declarations (and also
\AgdaKeyword{data}-declarations, but I use \AgdaKeyword{record}-declarations for
the convenience reason given in the previous paragraph).
The first feature is that the universe level of a record type has a lower bound
(the level of each field) but no upper bound.
Therefore, we can introduce the following declaration \AgdaRecord{Lift}, which
takes a type \AgdaBound{A} at level \AgdaBound{a} and produces an equivalent
type at a potentially higher level
\AgdaBound{a}\AgdaSpace{}\AgdaOperator{\AgdaPrimitive{$\sqcup$}}\AgdaSpace{}%
\AgdaBound{$\ell$}.
This type former is useful in situations which require a type at a specific
level, such as when constructing a type using a function.

### 2023-07-25 tex/conc.tex
[ writing ] full draft

\paragraph{Polymorphism}
An important feature of most contemporary statically typed programming languages
is polymorphism.
In particular, parametric polymorphism over types can be used to significantly
improve code reuse, and is well undestood theoretically via System $F$ and its
variants.
I have not considered polymorphism in this thesis, and neither did
\citet{AACMM21} in their paper, so whether it can be supported in the framework
presented earlier is an open question.
However, both \citet{Autosubst15} and \citet{Autosubst18} have applied similar
work to polymorphic calculi via special support, suggesting that it would be
possible to modify the framework of this thesis in a similar way.

>>>>

\paragraph{Polymorphism}
An important feature of most contemporary statically typed programming languages
is polymorphism.
In particular, parametric polymorphism over types can be used to significantly
improve code reuse, and is well understood theoretically via System $F$ and its
variants.
I have not considered polymorphism in this thesis, and neither did
\citet{AACMM21} in their paper, so whether it can be supported in the framework
presented earlier is an open question.
However, both \citet{Autosubst15} and \citet{Autosubst18} have applied similar
work to polymorphic calculi via special support, suggesting that it would be
possible to modify the framework of this thesis in a similar way.

### 2023-07-25 tex/examples.tex
[ writing ] full draft

In this chapter, I provide four example uses of semantic traversals:
a usage elaborator, a normalisation by evaluation algorithm, a denotational
semantics, and translations between calculi.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
The usage elaborator (\cref{sec:usage-elaborator}) gives an example of a program
that is generic in both syntax and usage annotations, as well as being an
essential tool for practical use of the framework.
The normalisation by evaluation algorithm (\cref{sec:nbe}) is for a specific
syntax, but exercises parts of the generic semantic traversal --- particularly,
its handling of environments --- to good effect.
The denotational semantics in terms of world-indexed relations
(\cref{sec:den-sem}) presents an interesting semantics for semiring-annotated
type systems, but with minimal use of operations provided by the framework.
Finally, the translations between $\name$ and L/nL (\cref{sec:lnl}) both justify
the encoding of L/nL given in \cref{sec:dup-lnl} and give examples of syntactic
work benefitting from the generic renaming and substitution operations.

@@

The type of syntax descriptions depends on the type of usage annotations because
of variable binding.
For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-bunched}, the right
premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
from the ambient posemiring.
The scaling combinator also makes direct reference to the posemiring.
To produce a simplified syntax description, where usage constraints are
trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
posemiring.
In contrast to syntax descriptions, even though types can contain usage
annotations, the type of types does not depend on the type of usage annotations.
This means that, in our simplified syntax, terms have types from the original
system even though variables have trivial usage annotations.
We define the $\mathbf0$ posemiring as follows, being careful to use the
0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
solved by Agda's $\eta$-laws.
Indeed, in this very definition, all of the semiring operations and laws are
canonically inferred.

>>>>

In this chapter, I provide four example uses of semantic traversals:
a usage elaborator, a normalisation by evaluation algorithm, a denotational
semantics, and translations between calculi.
The reader is also encouraged to see the far greater range of examples in the
work of \citet{AACMM21}, which should adapt to our usage-annotated setting.
The usage elaborator (\cref{sec:usage-elaborator}) gives an example of a program
that is generic in both syntax and usage annotations, as well as being an
essential tool for practical use of the framework.
The normalisation by evaluation algorithm (\cref{sec:nbe}) is for a specific
syntax, but exercises parts of the generic semantic traversal --- particularly,
its handling of environments --- to good effect.
The denotational semantics in terms of world-indexed relations
(\cref{sec:den-sem}) presents an interesting semantics for semiring-annotated
type systems, but with minimal use of operations provided by the framework.
Finally, the translations between $\name$ and L/nL (\cref{sec:lnl}) both justify
the encoding of L/nL given in \cref{sec:dup-lnl} and give examples of syntactic
work benefiting from the generic renaming and substitution operations.

@@

The type of syntax descriptions depends on the type of usage annotations because
of variable binding.
For example, in the $\oc{\gr r}$-E rule of \cref{fig:lr-bunched}, the right
premise binds a new variable with annotation $\gr r$, where $\gr r$ is drawn
from the ambient posemiring.
The scaling connective also makes direct reference to the posemiring.
To produce a simplified syntax description, where usage constraints are
trivialised, we set the ambient posemiring to the 1-element $\mathbf0$
posemiring.
In contrast to syntax descriptions, even though types can contain usage
annotations, the type of types does not depend on the type of usage annotations.
This means that, in our simplified syntax, terms have types from the original
system even though variables have trivial usage annotations.
We define the $\mathbf0$ posemiring as follows, being careful to use the
0-field record type \AgdaRecord{$\top$} so that everything algebraic gets
solved by Agda's $\eta$-laws.
Indeed, in this very definition, all of the semiring operations and laws are
canonically inferred.

### 2023-07-25 tex/framework.tex
[ writing ] expand and reorganise examples
[ writing ] full draft

In this chapter, I take the framework for typed syntaxes with binding developed
by \citet{AACMM21} and apply the principles we discovered in
\cref{sec:semirings} to yield a framework for defining calculi with
semiring-based usage restrictions on variables.
Syntactically, I claim that this framework ranges over all finitiary
variable-based simply typed semiring-annotated calculi, with justification by
comparison to the framework of \citet{AACMM21} and some novel examples in
\cref{sec:other-syntaxes}.
%I also derive analogues to some of the semantic results of \citet{AACMM21},
%strengthening them to take advantage of usage restrictions.

@@

\section{Example semantics}\label{sec:example-semantics}

>>>>

In this chapter, I take the framework for typed syntaxes with binding developed
by \citet{AACMM21} and apply the principles we discovered in
\cref{sec:semirings} to yield a framework for defining calculi with
semiring-based usage restrictions on variables.
Syntactically, I claim that this framework ranges over all finitary
variable-based simply typed semiring-annotated calculi, with justification by
comparison to the framework of \citet{AACMM21} and some novel examples in
\cref{sec:other-syntaxes}.
%I also derive analogues to some of the semantic results of \citet{AACMM21},
%strengthening them to take advantage of usage restrictions.

@@

\chapter{Applications}\label{sec:example-semantics}

### 2023-07-25 tex/gen-sem.tex
[ writing ] full draft

With the aim of abstracting away from explicit contexts, bringing us closer to
natural deduction, we can use some new notation to rephrase these requirements.
We will work in \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
One of the basic connectives in this setting is the \emph{pointwise} arrow
\AgdaFunction{\_$\dotto$\_}, which acts in
\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
like the non-dependent arrow does in \AgdaPrimitive{Set}.
Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} into \AgdaPrimitive{Set} by using an implicit $\Pi$-type
to quantify over \emph{all} contexts.
Finally, at this stage, I introduce a modality \AgdaFunction{$\bigcirc$}
encapsulating the pattern of considering arbitrary \emph{extensions} of a
context.
To facilitate working in this point-free setting, I give infix versions of
the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
\AgdaFunction{\_$\V\vDash$\_} and \AgdaFunction{\_$\C\vDash$\_}).
The principal use of these aliases is to fill the right argument with a type
(occuring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
a context given through the point-free machinery.

>>>>

With the aim of abstracting away from explicit contexts, bringing us closer to
natural deduction, we can use some new notation to rephrase these requirements.
We will work in \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} rather than \AgdaPrimitive{Set}.
One of the basic connectives in this setting is the \emph{pointwise} arrow
\AgdaFunction{\_$\dotto$\_}, which acts in
\AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}\AgdaPrimitive{Set}
like the non-dependent arrow does in \AgdaPrimitive{Set}.
Another basic component is the \AgdaFunction{$\forall[\_]$} notation, which
embeds \AgdaRecord{Ctx}\AgdaSpace{}\AgdaSymbol{$\to$}\AgdaSpace{}%
\AgdaPrimitive{Set} into \AgdaPrimitive{Set} by using an implicit $\Pi$-type
to quantify over \emph{all} contexts.
Finally, at this stage, I introduce a modality \AgdaFunction{$\bigcirc$}
encapsulating the pattern of considering arbitrary \emph{extensions} of a
context.
To facilitate working in this point-free setting, I give infix versions of
the families \AgdaBound{$\V$} and \AgdaBound{$\C$} (respectively
\AgdaFunction{\_$\V\vDash$\_} and \AgdaFunction{\_$\C\vDash$\_}).
The principal use of these aliases is to fill the right argument with a type
(occurring explicitly), and leave the left argument as \AgdaFunction{\_}, i.e.,
a context given through the point-free machinery.

### 2023-07-25 tex/gen-syn.tex
[ writing ] full draft

I will present a scheme based on the work of \citet{AACMM21} such that
\cref{fig:app-lam} is interpreted as the type
system we studied in the previous sections (simply typed $\lambda$-caluclus with
a base type and function types).
Remember that, while these look like inference rules, I am treating them
entirely formally, collected together into a \emph{syntax description}.
The information presented in \cref{fig:app-lam} is essentially all of the
information needed for the type system sans any details about variables.
In particular, notice:
\begin{itemize}
  \item Contexts, in particular the context of a rule's conclusion, which is
    shared in all premises in the resulting type system, are elided.
    The only part of any context I record is the newly bound variables in
    premises, such as the variable bound by a $\lambda$-abstraction.
  \item There is no explicit variable rule.
    It is understood that any $x : A$ in the context of the resulting type
    system can be used to yield a term with type $A$.
\end{itemize}

>>>>

I will present a scheme based on the work of \citet{AACMM21} such that
\cref{fig:app-lam} is interpreted as the type
system we studied in the previous sections (simply typed $\lambda$-calculus with
a base type and function types).
Remember that, while these look like inference rules, I am treating them
entirely formally, collected together into a \emph{syntax description}.
The information presented in \cref{fig:app-lam} is essentially all of the
information needed for the type system sans any details about variables.
In particular, notice:
\begin{itemize}
  \item Contexts, in particular the context of a rule's conclusion, which is
    shared in all premises in the resulting type system, are elided.
    The only part of any context I record is the newly bound variables in
    premises, such as the variable bound by a $\lambda$-abstraction.
  \item There is no explicit variable rule.
    It is understood that any $x : A$ in the context of the resulting type
    system can be used to yield a term with type $A$.
\end{itemize}

### 2023-07-25 tex/ill.tex
[ writing ] full draft

To encode what it means to use a hypothesis \emph{exactly once}, we first need
to decide what counts as a use.
The simplest case is that the identity sequent counts as a single use of its
sole hypothesis, and conversely does not count as a use of any other hypotheses.
For sequential proofs, created by the \TirName{Cut} rule, if we have a proof of
$A$ using $\Gamma$, and a proof of $B$ using $\Delta$ and $A$, then we have a
proof of $B$ transitively using $\Delta$ and $\Gamma$.
The exchange rule Exch says that use is invariant under permutation.

@@

Additional to the problems with \TirName{Promotion} in a natural deduction
system, defining the $\oc$-modality as in \cref{fig:bang-seq} has the odd
feature that, unlike all of the other connectives of ILL, $\oc$ is not
characterised by a universal property.
This can be seen by the fact that taking the rules for $\oc$ and replacing each
occurrence of $\oc$ by a fresh connective $\oc'$ produces a logically distinct
connective.
One cannot produce any derivation of $\oc' A \vdash \oc A$ because
\TirName{Promotion} does not apply when there are antecedents not of the form
$\oc X$.
This lack of characterisation also holds of the rule and axioms we first gave to
the $\Box$-modality in \cref{fig:S4-axioms}, and essentially any presentation
that does not encorporate the modality into the judgemental structure of the
calculus.

@@

The \citeauthor{BBdePH93} \TirName{Promotion} rule is a clever solution to the
substitution problem of (intuitionistic) linear logic, but the resulting system
comes with its own problems.
If we were to write programs in a language directly implementing the
\citeauthor{BBdePH93} calculus, then the \TirName{Promotion} rule would be a
pain point because it makes us rebind all of the $oc$-typed variables we need to
new variables.
Additionally, we have to be explicit everywhere about weakening and contraction,
potentially making working with $\oc$-typed variables much more fiddly than the
corresponding variables would be in a non-substructural programming language.
These syntactic annoyances may be worked around in a realistic implementation by
some new elaboration procedures, but we would prefer not to rely on such
elaboration if there is an acceptable core calculus which more closely matches
the programs we want to write.
Therefore, I investigate such a system, Dual Intuitionistic Linear Logic, in the
following subsection.

>>>>

To encode what it means to use a hypothesis \emph{exactly once}, we first need
to decide what counts as a use.
The simplest case is that the identity sequent counts as a single use of its
sole hypothesis, and conversely does not count as a use of any other hypotheses.
For sequential proofs, created by the \TirName{Cut} rule, if we have a proof of
$A$ using $\Gamma$, and a proof of $B$ using $\Delta$ and $A$, then we have a
proof of $B$ transitively using $\Delta$ and $\Gamma$.
The exchange rule \TirName{Exch} says that use is invariant under permutation.

@@

Additional to the problems with \TirName{Promotion} in a natural deduction
system, defining the $\oc$-modality as in \cref{fig:bang-seq} has the odd
feature that, unlike all of the other connectives of ILL, $\oc$ is not
characterised by a universal property.
This can be seen by the fact that taking the rules for $\oc$ and replacing each
occurrence of $\oc$ by a fresh connective $\oc'$ produces a logically distinct
connective.
One cannot produce any derivation of $\oc' A \vdash \oc A$ because
\TirName{Promotion} does not apply when there are antecedents not of the form
$\oc X$.
This lack of characterisation also holds of the rule and axioms we first gave to
the $\Box$-modality in \cref{fig:S4-axioms}, and essentially any presentation
that does not incorporate the modality into the judgemental structure of the
calculus.

@@

The \citeauthor{BBdePH93} \TirName{Promotion} rule is a clever solution to the
substitution problem of (intuitionistic) linear logic, but the resulting system
comes with its own problems.
If we were to write programs in a language directly implementing the
\citeauthor{BBdePH93} calculus, then the \TirName{Promotion} rule would be a
pain point because it makes us rebind all of the $\oc$-typed variables we need to
new variables.
Additionally, we have to be explicit everywhere about weakening and contraction,
potentially making working with $\oc$-typed variables much more fiddly than the
corresponding variables would be in a non-substructural programming language.
These syntactic annoyances may be worked around in a realistic implementation by
some new elaboration procedures, but we would prefer not to rely on such
elaboration if there is an acceptable core calculus which more closely matches
the programs we want to write.
Therefore, I investigate such a system, Dual Intuitionistic Linear Logic, in the
following subsection.

### 2023-07-25 tex/intro.tex
[ writing ] full draft

Within programming language research, type theory is a methodology for producing
abstractions via classifying program behaviour.
The simplest and most common type systems classify programs by what values they
may return.
For example, the body of a C function with return type \texttt{char} is a
program that, if it returns a value, will return a character.
If that function has a parameter of type \texttt{int}, then we can compose it
with a function with return type \texttt{int} to build up a larger program.
We expect type systems to stop us from running programs with arguments of the
wrong types.
This holds of both static and dynamic type systems --- with a static type
system, the compiler will refuse to compile our code if we pass a \texttt{char}
value to a function expecting an \texttt{int}, while with a dynamic type system,
our program will do a check before running the function to make sure that we
have passed it an \texttt{int}.
The abstraction produced in such simple type systems is the idea that a function
will take arguments in accordance with its parameter types and produce a result
in accordance with its result type, and furthermore, as readers, we do not have
to inspect the function's implementation to see that these properties are true.

@@

The work of this thesis relies upon type theory in two distinct ways.
Firstly, as I have introduced above, the main objects of study in this thesis
are programming languages with interesting type systems.
Secondly, however, type theory provides the basis of the proof assistant Agda I
use to implement the aforementioned programming languages and operations upon
them.
I will now introduce the idea of proof assistants.

@@

Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
this thesis --- double as programming languages themselves.
This means that we can write progams and prove properties of them using the same
tool.
Also, many theorems proven in such a proof assistant have computational content.
For example, if we prove a normalisation theorem for a programming language,
this will typically yield a (verified) normalisation algorithm for it, which we
can really run on a computer.
As such, a development in a proof assistant can provide a reference
implementation of a programming language, or even --- as with Idris
2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
actual implementation of a programming language.

@@

I assume familiarity with the Curry-Howard correspondence throughout this
thesis.
I make no distinction between logics and type theories, and use terminology from
each interchangeably.
Each following bullet point lists a collection of synonyms.

\begin{itemize}
  \item assumption, hypothesis, variable
  \item derivation, proof, term
  \item proposition, formula, type
  \item connective, type former
\end{itemize}

>>>>

Within programming language research, type theory is a methodology for enforcing
abstractions by classifying program behaviour.
The simplest and most common type systems classify programs by what values they
may return.
For example, the body of a C function with return type \texttt{char} is a
program that, if it returns a value, will return a character.
If that function has a parameter of type \texttt{int}, then we can compose it
with a function with return type \texttt{int} to build up a larger program.
We expect type systems to stop us from running programs with arguments of the
wrong types.
This holds of both static and dynamic type systems --- with a static type
system, the compiler will refuse to compile our code if we pass a \texttt{char}
value to a function expecting an \texttt{int}, while with a dynamic type system,
our program will do a check before running the function to make sure that we
have passed it an \texttt{int}.
The abstraction produced in such simple type systems is the idea that a function
will take arguments in accordance with its parameter types and produce a result
in accordance with its result type, and furthermore, as readers, we do not have
to inspect the function's implementation to see that these properties are true.

@@

The work of this thesis relies upon type theory in two distinct ways.
Firstly, as I have introduced above, the main objects of study in this thesis
are programming languages with interesting type systems.
Secondly, type theory provides the basis of the proof assistant Agda I
use to implement the aforementioned programming languages and operations upon
them.
I will now introduce the idea of proof assistants.

@@

Finally, many proof assistants --- including Agda~\citep{Agda}, which I use in
this thesis --- double as programming languages themselves.
This means that we can write programs and prove properties of them using the same
tool.
Also, many theorems proven in such a proof assistant have computational content.
For example, if we prove a normalisation theorem for a programming language,
this will typically yield a (verified) normalisation algorithm for it, which we
can really run on a computer.
As such, a development in a proof assistant can provide a reference
implementation of a programming language, or even --- as with Idris
2~\citep{Brady21}, Lean 4~\citep{deMU21}, and Cedille~\citep{GRS16} --- the
actual implementation of a programming language.

@@

I assume familiarity with the Curry-Howard correspondence~\citep{Howard80}
throughout this thesis.
I make no distinction between logics and type theories, and use terminology from
each interchangeably.
Each following bullet point lists a collection of synonyms.

\begin{itemize}
  \item assumption, hypothesis, variable
  \item proposition, formula, type
  \item connective, type former
  \item derivation, proof, term
  \item derivable (formula), inhabited (type)
\end{itemize}

### 2023-07-25 tex/kits.tex
[ writing ] full draft

In the derivation below, we assume the existence of a derivation of
$B, C \to C \vdash C$, and by the admissibility of substitution we thus have a
derivation of $A \to B, A \vdash C$.
Intuitively, the context $A \to B, A$ derives the context $B, C \to C$, so
anything derived from $B, C \to C$ can also be derived from $A \to B, A$.
We see formally that $A \to B, A$ derives $B, C \to C$ by deriving each element
of the latter from the former --- hence the first two premises of the Subst
rule below, deriving $B$ and $C \to C$ from $A \to B, A$.

>>>>

In the derivation below, we assume the existence of a derivation of
$B, C \to C \vdash C$, and by the admissibility of substitution we thus have a
derivation of $A \to B, A \vdash C$.
Intuitively, the context $A \to B, A$ derives the context $B, C \to C$, so
anything derived from $B, C \to C$ can also be derived from $A \to B, A$.
We see formally that $A \to B, A$ derives $B, C \to C$ by deriving each element
of the latter from the former --- hence the first two premises of the \TirName{Subst}
rule below, deriving $B$ and $C \to C$ from $A \to B, A$.

### 2023-07-25 tex/linmech.tex
[ writing ] full draft

The original paper extends the logic covered to binary additives --- $\with$ and
$\oplus$ --- with rules that check that terms agree on output contexts, as seen
in \cref{fig:twl-add}.
However, it is less clear how to handle nullary additives --- $\top$ and $0$ ---
as we would have 0 (rather than 2) potential candidates for the output context.
At some level, this problem is unavoidable in a system modelling linearity
checking because any checking strategy will expose the ambiguity in sequents
like $\gr1x : A \vdash (\langle\rangle, \langle\rangle) : \top \otimes \top$ of
whether the variable
$x$ was consumed in the left half or the right half.
Such an example is also considered in related work on proof seach for linear
logics, such as the work of \citet[p.\ 11]{WH94} and \citet[p.\ 150]{CHP00}.
It is not immediately clear whether the different solutions proposed by these
papers will apply to \citeauthor{allais17}' and my settings, given that they
both act on a set of formulas restricted to facilitate proof search.
The solutions also add significant, seemingly somewhat ad hoc, structure to the
syntax of sequents, with no semantic justification (rather being justified by
making their respective proof search algorithms efficient).
%As such, we may expect uses of $\top$-introduction (and similarly
%$0$-elimination) to be annotated with which variables they consume, in which case we
%get a viable typing with leftovers rule.

>>>>

The original paper extends the logic covered to binary additives --- $\with$ and
$\oplus$ --- with rules that check that terms agree on output contexts, as seen
in \cref{fig:twl-add}.
However, it is less clear how to handle nullary additives --- $\top$ and $0$ ---
as we would have 0 (rather than 2) potential candidates for the output context.
At some level, this problem is unavoidable in a system modelling linearity
checking because any checking strategy will expose the ambiguity in sequents
like $\gr1x : A \vdash (\langle\rangle, \langle\rangle) : \top \otimes \top$ of
whether the variable
$x$ was consumed in the left half or the right half.
Such an example is also considered in related work on proof search for linear
logics, such as the work of \citet[p.\ 11]{WH94} and \citet[p.\ 150]{CHP00}.
It is not immediately clear whether the different solutions proposed by these
papers will apply to \citeauthor{allais17}' and my settings, given that they
both act on a set of formulas restricted to facilitate proof search.
The solutions also add significant, seemingly somewhat ad hoc, structure to the
syntax of sequents, with no semantic justification (rather being justified by
making their respective proof search algorithms efficient).
%As such, we may expect uses of $\top$-introduction (and similarly
%$0$-elimination) to be annotated with which variables they consume, in which case we
%get a viable typing with leftovers rule.

### 2023-07-25 tex/lnd.tex
[ add ] Restall reference
[ writing ] full draft

\subsection{$\name$ stated using bunched connectives}\label{sec:lr-bunched}
The full system \name{} is stated in terms of bunched connectives in
\cref{fig:lr-bunched}.
The bunched connectives also yield a reasonably consise definition of the Agda
data type of \name{} derivations, as seen in \cref{fig:lr-bunched-Agda}.

@@

The separating conjunction $\sep$ can be seen as a decategorified version of Day
convolution~\citep{Day70}.
It also resembles the use of ternary frames \todo{cite} in semantics of
substructural logics.

>>>>

\subsection{$\name$ stated using bunched connectives}\label{sec:lr-bunched}
The full system \name{} is stated in terms of bunched connectives in
\cref{fig:lr-bunched}.
The bunched connectives also yield a reasonably concise definition of the Agda
data type of \name{} derivations, as seen in \cref{fig:lr-bunched-Agda}.

@@

The separating conjunction $\sep$ can be seen as a decategorified version of Day
convolution~\citep{Day70}.
It also resembles the use of ternary frames in semantics of non-distributive
logics~\citep[chapter 12]{Restall99}.
